Data Analytics Introduction Using Python Training

### 2024-11-11 MsaPythonDataAnalytics.qmd
'day1'

@@

### Iterations

@@

@@

@@

@@

>>>>

```{python}
#| echo: true
#| eval: false
print(first_string.capitalize())
print(first_string.title())
print(first_string.upper())
print(first_string.swapcase())
print(first_string.find('is'))
print(first_string.replace('first', 'second'))
print(first_string.strip())
print(first_string.isalnum())
print(first_string.isalpha())
print(first_string.isdigit())
print(first_string.isprintable())
```

@@

### Conditions & Iterations

Conditions criteria where we compare values and decide what step to take.
Example of conditional criteria `if-else` , `if-elif-else`, `while`.

if condition:
  #do something
else:
  #do something
```

Comparison operators go hand in hand with conditions.
Comparison operators inculde `==` , `<=`, `>=`, `|`, `&`, `or`, `and` e.t.c

@@

### Date & Time

- **Date:** Handles dates without time.
- **POSIXct:** Handles date & time in calendar time.
- **POSIXlt:** Handles date & time in local time.
- **Hms:** Parses periods with hour, minute, and second
- **Timestamp:** Represents a single pandas date & time
- **Interval:** Defines an open or closed range between dates and times
- **Time delta:** Computes time difference between different datetimes

***ISO8601 datetime format***  (YYYY-MM-DD HH:MM:SS TZ)

```{python}
#| echo: true
#| eval: true

import datetime as dt
import time as tm
import pytz
import pandas as pd

```{python}
#| echo: true
#| eval: true

# Get the current date
dt.date.today()

# Get the current date and time
dt.datetime.now()
```

***Parsing dates, datetimes, and times***

```{python}
#| echo: true
#| eval: true

iso = pd.to_datetime(['2018-10-26 12:00', '2018-10-26 13:00'], utc=True)

us = ['07/20/1969 20:17:40','11/19/1969 06:54:35']

non_us = ['20/07/1969 20:17:40','19/11/1969 06:54:35']

parts = pd.DataFrame({
  'year' : [1969,1969,1971],
  'month' : [7,11,2],
  'day' : [20,19,5]
})

# Parse dates in ISO format
pd.to_datetime(iso)

# Parse dates in US format
pd.to_datetime(us, dayfirst=False)

# Parse dates in NON US format
pd.to_datetime(non_us, dayfirst=True)

# Parse dates, guessing a single format
pd.to_datetime(iso, infer_datetime_format=True)

# Parse dates in single, specified format
pd.to_datetime(iso, format="%Y-%m-%d %H:%M:%S")

# Parse dates in single, specified format
pd.to_datetime(us, format="%m/%d/%Y %H:%M:%S")

# Make dates from components
pd.to_datetime(parts)
```

***Extracting components***

```{python}
#| echo: true
#| eval: true

# Parse strings to datetimes
dttm = pd.to_datetime(iso)

# Get year from datetime pandas series
dttm.year

# Get day of the year from datetime pandas series
dttm.day_of_year

# Get month name from datetime pandas series
dttm.month_name()

# Get day name from datetime pandas series
dttm.day_name()

# Get datetime.datetime format from datetime pandas series
dttm.to_pydatetime()
```

**Rounding Dates**

```{python}
#| echo: true
#| eval: true

# Rounding dates to nearest time unit
dttm.round('1min')

# Flooring dates to nearest time unit
dttm.floor('1min')

# Ceiling dates to nearest time unit
dttm.ceil('1min')
```

**Arithmetic**

```{python}
#| echo: true
#| eval: true

# Create two datetimes
now = dt.datetime.now()
then = pd.Timestamp('2021-09-15 10:03:30')

# Get time elapsed as timedelta object
now - then

# Get time elapsed in seconds 
(now - then).total_seconds()

# Adding a day to a datetime
dt.datetime(2022,8,5,11,13,50) + dt.timedelta(days=1)
```

**Time Intervals**

```{python}
#| echo: true
#| eval: true

# Create interval datetimes
start_1 = pd.Timestamp('2021-10-21 03:02:10')
finish_1 = pd.Timestamp('2022-09-15 10:03:30')
start_2 = pd.Timestamp('2022-08-21 03:02:10')
finish_2 = pd.Timestamp('2022-12-15 10:03:30')

# Specify the interval between two datetimes
pd.Interval(start_1, finish_1, closed='right')

# Get the length of an interval
pd.Interval(start_1, finish_1, closed='right').length

# Determine if two intervals are intersecting
pd.Interval(start_1, finish_1, closed='right').overlaps(pd.Interval(start_2, finish_2, closed='right'))
```

**Time Deltas**

```{python}
#| echo: true
#| eval: true

# Define a timedelta in days
pd.Timedelta(7, "d")

# Convert timedelta to seconds
pd.Timedelta(7, "d").total_seconds()
```

## Joins & Merge {visibility="hidden"}

@@

Python provides so many modules for visualization but for purposes of the training we will deal with the two most common visualiation tools.

1. **matplotlib**
1. **seaborn**
1. plotnine
1. altair
1. Bokeh

@@

**Box plots**

# Sample data
data = {
    'Category': ['A']*10 + ['B']*10,
    'Value': [1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11]
}

df = pd.DataFrame(data)

# Box plot
df.boxplot(by='Category')
plt.title('Box Plot Example')
plt.suptitle('')
plt.xlabel('Category')
plt.ylabel('Value')
plt.show()

@@

#import seaborn as sns

tips = sns.load_dataset("tips")

sns.scatterplot(x="total_bill", y="tip", data=tips)
```

import seaborn as sns

import matplotlib.pyplot as plt

tips = sns.load_dataset("tips")

# customize the scatter plot

sns.scatterplot(x="total_bill", y="tip", hue="sex", size="size", sizes=(50, 200), data=tips)

# add labels and title

plt.xlabel("Total Bill")

plt.ylabel("Tip")

plt.title("Relationship between Total Bill and Tip")

# display the plot

plt.show()
```

### Pandas Inbuilt Visualization

* df.plot.area     
* df.plot.barh     
* df.plot.density  
* df.plot.hist     
* df.plot.line     
* df.plot.scatter
* df.plot.bar      
* df.plot.box      
* df.plot.hexbin   
* df.plot.kde      
* df.plot.pie
